Predicting Remaining Useful Life of NASA Aircraft Engines

### 2025-03-24 index.qmd
fixed errors

where $S(x)$ is the probability that an individual has survived past
time $x$ [@klein2005]. undefined

@@

The function is used to describe the changes in [@klein2005].

Since CPH is a proportional hazard model, the hazard function is adapted
to deal with the relationship between time to event and the explanatory
variables [@klein2005]. It can be defined as

@@

### Data Structure

In order to use CPH to model time to event survival data, the data must
contain a value for time, a binary field indicating the event being
modeled, an id field, and one or multiple covariate fields that can be
categorical or numerical. The time value is typically formatted as a
time series for each subject in the dataset. The first entry for
subjects will be a time value from the beginning of the observation
period with the final entry time varying by subject based on if the
event occurred for that individual or if the study ended before the
event occurred. Subjects who did not have the event occur by the end of
the observation period are considered censored entries and will be
handleded different in the model [@nahhas2025].

@@

CPH assumes the relationship between covariates and the outcome is a
linear relationship [@nahhas2025]. For CPH, the outcome for that
relationship is the log of the hazard rate. This assumption is checked
visually using the Martingale residuals for each covariate
[@nahhas2025]. Martingale residuals are a type of residauls used in
survival analysis that show the discrepancy between the observed and the
expected number of events. The equation is
$\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}$.
The resulting residuals are examined visually and if they are linear and
appear to have a slope of zero, the linearity assumption is not violated
[@Amini2015].

Below is an example plot showing Martingale residuals plotted for a
covariate being tested. From the plot, it is visually clear that the
residuals are linear and appear to have a slope of zero. For this
covariate then, the linearity assumption has not been violated. If a
model is being generated with multiple covariates, multiple Martingale
residual plots will need to be generated in order to verify this
assumption for each covariate.

@@

the `cox.zph()` calculates the scaled Schoenfeld partial residuals and
returns the $\chi^2$ value, the degrees of freedom, and p-values for
each individual covariate along with a global value. Using an
$\alpha = 0.05$ and the global p-value, the determination of whether to
reject or fail to reject $H_0$ can be made. In addition to the
hypothesis test, these Shoenfeld residuals can be plotted against time
for each covariate to visualize the assumption. When plotting the
residuals against time, if the assumption is not broken it would be
expected that the points show roughly a straight horizontal line to
indicate constant hazard over time [@nahhas2025].

@@

@@

summary(cox)
```

#### Checking Assumptions

**Checking Proportional Hazards Assumption**

```{r}
cox.zph(cox)

plot(cox.zph(cox))

@@

**Checking Linearity Assumption**

Example for one predictor

```{r}
martingale <- resid(cox, type = "martingale")

@@

```{r}
survfit <- survfit(cox)

>>>>

where $S(x)$ is the probability that an individual has survived past
time $x$ [@klein2005].

@@

The function is used to describe how the chance of experiencing the
event changes with time and only has the requirement of being
non-negative [@klein2005].

CPH then uses both functions to develop it's own model, specifically by
adapting the hazard function. Since CPH is a proportional hazard model,
the hazard function is adapted to deal with the relationship between
time to event and the explanatory variables [@klein2005]. It can be
defined as

@@

@@

CPH assumes the relationship between covariates and the outcome is a
linear relationship [@nahhas2025]. For CPH, the outcome for that
relationship is the log of the hazard rate. This assumption is checked
visually using the Martingale residuals for each covariate
[@nahhas2025]. Martingale residuals are a type of residauls used in
survival analysis that show the discrepancy between the observed and the
expected number of events. The equation is $$
\text{Martingale Residuals} = \text{Observed Events} - \text{Expected Events}
$$ The resulting residuals are examined visually and if they are linear
and appear to have a slope of zero, the linearity assumption is not
violated [@Amini2015]. If a model is being generated with multiple
covariates, multiple Martingale residual plots will need to be generated
in order to verify this assumption for each covariate.

Below is an example plot showing Martingale residuals plotted for a
covariate being tested. From the plot, it is visually clear that the
residuals are linear and appear to have a slope of zero. For this
covariate then, the linearity assumption has not been violated.

@@

The scaled Schoenfeld partial residuals are calcuated and returns the
$\chi^2$ value, the degrees of freedom, and p-values for each individual
covariate along with a global value. Using an $\alpha = 0.05$ and the
global p-value, the determination of whether to reject or fail to reject
$H_0$ can be made. In addition to the hypothesis test, these Shoenfeld
residuals can be plotted against time for each covariate to visualize
the assumption. When plotting the residuals against time, if the
assumption is not broken it would be expected that the points show
roughly a straight horizontal line to indicate constant hazard over time
[@nahhas2025].

@@

### Evaluation

@@

cox1 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox2 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5)

cox3 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm5 + sm6 + sm7 + sm13 + sm14 + sm19 + sm21)

cox4 <- coxph(data = train, 
             Surv(time, status) ~ sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)

cox5 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19 + sm20 )

cox6 <- coxph(data = train, 
             Surv(time, status) ~ os1  + sm1 + sm2 + sm5  + sm7 + sm11 + sm12 + sm13 + sm19  + sm21)

cox7 <- coxph(data = train, 
             Surv(time, status) ~ os1  + sm13 + sm14 + sm19 + sm20)

cox8 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm21)

cox9 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm2 + sm5 + sm6 + sm12 + sm13  + sm19 + sm20 + sm21)

cox10 <- coxph(data = train, 
             Surv(time, status) ~  os3 + sm1 + sm2 + sm5 + sm6  + sm11 + sm12 + sm13  + sm19 + sm20 + sm21)

@@

The resulting models and their corresponding concordance index values
are below.

```{r}
concordance_results <- data.frame(row.names = c(
  "Model 1",
  "Model 2",
  "Model 3",
  "Model 4",
  "Model 5",
  "Model 6",
  "Model 7",
  "Model 8",
  "Model 9",
  "Model 10"),
  Concordance = c(
  cox1$concordance[6],
  cox2$concordance[6],
  cox3$concordance[6],
  cox4$concordance[6],
  cox5$concordance[6],
  cox6$concordance[6],
  cox7$concordance[6],
  cox8$concordance[6],
  cox9$concordance[6],
  cox10$concordance[6]))

kable(concordance_results)

The analysis will proceed using model 1 because of the high concordance
index. The full summary of model 1 is below.

```{r}
summary(cox1)

### Checking Assumptions

Despite the high concordance value for model 1, the assumptions will
need to be evaluated before using the model to make predictions. If any
of the assumptions for the model are not true, the model will need to be
modified

#### Checking Proportional Hazards Assumption

The `cox.zph` function can be used to check the proportional hazards
assumption. As mentioned above, the function calculates p values for
each covariate. If the p values are \< .05, the null hypothesis is
rejected indicating the proportional hazards assumption is true. If the
proportional hazards assumption is not true for any covariate, the model
will be recreated without those covariates.

```{r}
cox.zph(cox1)
```

The output above shows the p values for the hypothesis test.

In addition to the hypothesis test, the Shoenfeld residuals calculated
by the `cox.zph()` function are plotted against time below.

```{r}
# Function to put ph_plots on one visual
create_ph_plots <- function(coxmodel){
  # Function takes argument that is a cox model from coxph()

  phresults <- cox.zph(coxmodel)

  # Define dimensions of visual
  par(mfrow = c(3,5),
      mar = c(2,2,4,1))

  # Loop through each covariate creating a plot and adding it to the phplot list
  for (i in 1:(length(phresults$table[,1])-1)){
    phplots <- list()
    phplots[i] <- plot(phresults[i], # Create plot for each iteration
                       main = rownames(phresults$table)[i]) # Title of each plot is the rowname from coxmodel$table

  }
  mtext("Schoenfeld Residuals vs. Time for Covariates", side = 3, line = -1.5, outer = TRUE, cex = 1) # Title for Visual
  par(mfrow = c(1,1)) # Reset plotting area
}

create_ph_plots(cox1)
```

The proportional hazards test shows several covariates do not meet the
proportional hazards assumption.

The model is recreated below with another proportional hazards test.

```{r}

cox1.1 <- coxph(data = train, 
             Surv(time, status) ~ os1 + os3 + sm1 + sm2 + sm5 + sm6 + sm7 + sm11 + sm12 + sm13 + sm14 + sm19 + sm20 + sm21)
cox.zph(cox1.1)

The hypothesis test shows the covariates sm1, sm2, sm6, sm7, sm11, sm12,
sm20, and sm21 do not meet the proportional hazards assumption. The
model is created again below with a second hypothesis test showing all
of the covariates pass the hypothesis test.

#### Checking Linearity Assumption

Residuals can first be found by using the `resid(type = "martingale")`
function and passing martingale in the type argument creating a vector
of residuals that can be plotted against the range of values for each
covariate. Example for one predictor

```{r}
martingale <- resid(cox1.1, type = "martingale")

@@

The survival curve will be plotted by using the survival package's
`survfit()` function. The object created will be plotted using `plot()`

```{r}
survfit <- survfit(cox1.1)
